---
title: 'STA501 Week #3 Assignment'
author: 'Due: 11:30 PM, Sunday, 2/14/2021'
date: "2/9/2021"
output:
  word_document:
    toc: yes
    toc_depth: '4'
  html_document:
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    fig_caption: yes
    number_sections: yes
    theme: readable
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
---

<style type="text/css">
h1.title {
  font-size: 20px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 3 - and the author and data headers use this too  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}
h2 { /* Header 3 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}
h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 15px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}
h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}
</style>


```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
library(knitr)
knitr::opts_chunk$set(echo = TRUE,           # include code chunk in the output file
                      warnings = FALSE,       # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE          # you can also decide whether to include the output
                                              # in the output file.
                      )   
```


# Introduction and Data Source

This week's assignment focuses on the descriptive statistics using R. The **Diabetes** data set to be used in this assignment is taken from [Vanderbilt's Biostatistics Datasets](https://hbiostat.org/data/).

The following is the description from the web page:

These data are courtesy of Dr. John Schorling, Department of Medicine, University of Virginia School of Medicine.
The data consist of 19 variables on 403 subjects from 1046 subjects who were interviewed in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors in central Virginia for African Americans. According to Dr. John Hong, Diabetes Mellitus Type II (adult-onset diabetes) is associated most strongly with obesity. The waist/hip ratio may be a predictor of diabetes and heart disease. DM II is also associated with hypertension - they may both be part of "Syndrome X". The 403 subjects were the ones who were actually screened for diabetes. Glycosolated hemoglobin > 7.0 is usually taken as a positive diagnosis of diabetes. For more information about this study see

Willems JP, Saunders JT, DE Hunt, JB Schorling: Prevalence of coronary heart disease risk factors among rural blacks: A community-based study. *Southern Medical Journal* 90:814-820; 1997

Schorling JB, Roach J, Siegel M, Baturka N, Hunt DE, Guterbock TM, Stewart HL: A trial of church-based smoking cessation interventions for rural African Americans. *Preventive Medicine* 26:92-101; 1997.



```{r}
diabetes <- read.csv("https://raw.githubusercontent.com/pengdsci/STA501/main/Data/diabetes.csv", header = TRUE)
kable(head(diabetes))
```

We can see from the first 6 observations that there are 15 numerical variables and 3 categorical variables. Variable **bp.2s** and **bp.2d** have missing values. To complete this week's assignment, you need to choose one numerical variable and one categorical variable with NO missing values.

The following code shows how to extract variables from the data frame. I will use the two variables with missing values as an example. You can modify the code to extract your variables for the assignment.

```{r}
bp.2s <- diabetes$bp.2s
bp.2d <- diabetes$bp.2d
```

# Assignments for This Week: <font color = "blue"><b>Due: 11:30 PM, Sunday, 2/14/2021</b></font>

This assignment focuses on descriptive statistics on categorical and numerical data. Please prepare an R Markdown document to complete the assignment. <font color = "red">You are expected to submit the RMarkdown document and one of the three converted documents: HTML, PDF, and Word.</font>

## Summarizing Categorical Data

Use the categorical variable you selected to perform the following analysis

1. **Construct a <font color = "red">relative frequency table</font>**. write a few sentences to describe the distribution of the variable.  Note that you are encouraged to construct a frequency table with all four types of frequencies as I did in the class note. 

2. **Construct a pie-chart** to represent the distribution of the categorical variable.

## Summarizing Numerical Data

Using the numerical variable you chose from the diabetes data to answer the following questions.

3. **Construct a <font color = "red">relative frequency table of the numerical variable</font>** with 10 categories. In other words, the frequency table should have 10 rows. You are encouraged to include all 4 frequencies in the table. Please provide a brief description of the relative frequencies.


4. **Construct a histogram of the numerical variable** with 10 vertical bars. In other words, the histogram is a geometric representation of the frequency table. Explain the distribution of the variable. Is it skewed to the left or the right?

5. **Construct a box-plot** and explain it. That is, can you tell whether the distribution is skewed to the right or the left?  












