---
title: 'STA501 Week #02 Topics'
author: "Cheng Peng"
date: "01/30/2021"
output:
  html_document: 
    number_sections: yes
    toc: yes
    theme: cerulean
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    fig_caption: yes
  pdf_document:
    toc: yes
    toc_depth: 4
    fig_width: 5
    fig_height: 4
    fig_caption: yes
    number_sections: yes
    keep_tex: yes
  word_document:
    toc: yes
    toc_depth: '4'
---


<style type="text/css">
h1.title {
  font-size: 20px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 3 - and the author and data headers use this too  */
    font-size: 22px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}
h2 { /* Header 3 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 15px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}
</style>



```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
knitr::opts_chunk$set(echo = TRUE,            # include code chunk in the output file
                      warnings = TRUE,       # sometimes, you code may produce warning messages,
                                              # you can choose to include the warning messages in
                                              # the output file. 
                      results = TRUE          # you can also decide whether to include the output
                                              # in the output file.
                      )   
```


# Introduction

We have introduced the sampling plans and study designs to obtain valid data for statistical analysis.

## Sampling Plans

In general, probability sampling plans generate samples suitable for inferential statistics (such as confidence intervals and testing hypotheses).

### Simple Random Sampling (SRS)

In this sampling plan, each subject is randomly chosen in such a way that each subject in the population has an equal chance, or probability, of being selected. In mathematical statistics, there is a rigorous definition of SRS plan.


### Systematic sampling

In this sampling plan, we (randomly) label all subjects in a population from 1, 2, ..., N (population size), randomly choose a label, say $n_0$, and then select every $k^{th}$ subjects to include in the sample. That resulting sample is called a systematic sample. Because the 


### Stratified sampling

This sampling plan assume that the target population of interest has several naturally defined sub-populations, then we take a sub-sample (SRS) from each sub-population such that the sub-sample sizes are proportional to their corresponding sub-population sizes. 


## Study Designs

Depending on whether the variables in the study were modified or filtered by researchers, we can categorize the study designs into observational and experimental studies.

### Observational Studies

Observational studies are ones where researchers observe the effect of a risk factor, diagnostic test, treatment or other intervention **without trying to remove confounding factors**.

**Cross-sectional studies** collection information of a population by taking a snap-shot or cross-section of the population.  These studies usually involve one contact with the study population and are relatively cheap to undertake. 

**Pre-test/post-test** take two cross-sectional samples taken from a randomly selected subjects before and after an intervention (such as a new treatment) applied to the randomly selected subjects in the study. The objective is to see whether a characteristic of the sample changed before and after the intervention. 

**Retrospective studies** use the historical information to study the characteristic of a population. Sometimes current information of the population is not available or difficult to obtain, we then use the historical information **under certain assumptions**. 

**Longitudinal studies** follow study subjects over a period of time with repeated data collection throughout. Most are observational studies that seek to identify a correlation among various factors. Thus, longitudinal studies do not manipulate variables and are not often able to detect causal relationships.


### Experimental Designs

Experimental studies are ones where researchers **carefully design experiments to remove potential confounding factors** and introduce an intervention and then study the effects.

**Prospective studies**, also called **follow-up study**, in which you select a random select subjects and wait for a period of time to record the formation of interest to perform statistical analysis. 

**Randomized controlled trials (RCT)** in clinical studies are always prospective studies and often involve following a “cohort” of individuals to determine the relationship between various variables.

**Longitudinal studies** can also be considered as experimental study.


# Loading Data to R from External Data Files

Three basic types of data file are common in practice: csv, txt, and xls(x). I have uploaded the well-known **iris* data set to the course web page in the aforementioned formats. You can practice the R functions to load these external data sets to R.


## Text File ( aka. Delimited Text File)

R command **read.table()** will load text files in R. 

**Caution:** if the data file contains **missing values**, you need to handle the missing values before loading it to R. If you take a formal programming course, you will be write several line of codes to clean the data. In this class, we only use basic R commands to do analysis.

The code in the following code chunk does not work!!!

```{}
placement.data = read.table("C:\\STA501\\w02\\placement.txt", header=TRUE) 
```


### Read files from remote web server

```{r}
### delimited text file
iris.text = read.table("https://stat501.s3.amazonaws.com/w02-iris.txt", header = TRUE)
### csv file
iris.csv = read.csv("https://stat501.s3.amazonaws.com/w02-iris.csv", header = TRUE)
```

Note that there is no commands such as **read.table()** and **read.csv()** in the base R to read Excel file from URL. There are several R functions in different libraries, such as **read_xlsx()** and **read_xls()** in library **{readxl}**, can read Excel file from the local drive (see the example in the next sub-setion).


### Read files from local folder

```{r}
### delimited text file
iris.text.loc = read.table("C:\\STA501\\w02\\w02-iris.txt", header = TRUE)
### csv file
iris.csv.loc = read.table("C:\\STA501\\w02\\w02-iris.csv", header = TRUE)
## Excel file - no built-in command in the base R can read Excel file to R. 
## need to load a command in a R library.
library(readxl)
iris.xlsx.loc = read_xlsx("C:\\STA501\\w02\\w02-iris.xlsx")
```



# Data Frame and List

Data frame can hold different type of variables but requires variables to be equal in length. If you have variables in different types **and** having different length, you need to use list to hold these variables.


## Data Frame

* We define following categrical and numeric vectors (data sets)

```{r}
veca = 1:26           #  c() is unnecessary since colon (:) is a shortcut to
                      #  define a patterned sequence. 1, 2, 3, 4, ..., 24, 25, 26.
vecb = 101:126
vecc = 201:226
vecd = letters        #  26 letters (lower case)
vece = LETTERS        #  upper-case letters
```

* **Example 1**.  Define data frame with different types of variables

```{r}
dataframe01 = data.frame(A=veca, B = vecb, D = vecd, E = vece)  # 26-by-3 data frame
```

* **Example 2**. Define a data frame using vectors with unequal width. The issue is that R will recycle the small vectors **whose lengths are factors of the length of longest vector** to make equal lengths columns in the data frame. 

```{r}
## CAUTION: If the lengths of individual variables are different, R will recycle the values in the 
##          small data (short vector) to make the equal length across the columns
##
dframe2 = data.frame(A = veca,                  # this is a vector with 100 values
                     X = 1:13,                  # this vector has 13 values
                     c = sum(veca+vecb+vecc)    # two comments: 1. c is reserved for defining
                                                # vectors. Should not be used a name                                                                               # of any objects in R
                                                # 2. this is also a single value.
  )
```


* **Example 3**. Following code will produce error because Y has 4 rows and A has 26 rows!

```{r eval=FALSE}
dframe2 = data.frame(A = veca,                  # this is a vector with 100 values
                     Y = c(33,44,55,66),        #  4 values in vector b!
                     W = c(99,999)              #  2 values in W!
)
```
<font color = "red"> Error in data.frame(a = veca, b = c(33, 44, 55, 66), c = c(99, 999)) : arguments imply differing number of rows: 26, 4, 2 </font>


you can check whether the data frame is correctly defined. I added an option **eval = FALSE** to the following code chunk to avoid printing out the 100-by-3 data frame.

```{r results = FALSE}
dframe2
```


```{r caption = "This figure is meaningless since the variables in the data frame are not the original variables due to unequal number of rows leading to value recycling."  }
plot(dframe2)   # no error. This may not be what you expected since the data frame has recycling issues.
```



The following code does not work since no specific column is specified to calculate the standard deviation!



```{}
sd(dframe2)   # sd() calculates the standard deviation of a SINGLE vector, but
              # there 3 in the data frame. The error message says that R cannot
              # cannot combine all column to make a vector. In R, numerical 
              # values are 'double'.
```

you can calculate the standard deviation of the variables in the data frame one by one. For example,

```{r}
sd(dframe2$A)    # This will NOT generate error since you calculate the standard deviation for a specific variable (column).
```



# Taking Random Samples from a Target Population Using R

to be completed on Thursday.